---
title: "Optiver Volatility Prediction"
author: "Optiver 3"
date: "2024-05-14"
output: html_document
execute:
  cache: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r message=FALSE, warning=FALSE, show=FALSE}
# Function to install packages if not already installed
install_if_missing <- function(packages) {
  new_packages <- packages[!(packages %in% installed.packages()[,"Package"])]
  if(length(new_packages)) install.packages(new_packages, dependencies = TRUE)
  invisible(lapply(packages, library, character.only = TRUE))
}

# List of required packages
required_packages <- c("tidyverse", "rugarch", "data.table", "plotly", "cluster",
                       "class", "vars", "rmgarch", "parallel", "gridExtra", "factoextra")

# Install and load packages
install_if_missing(required_packages)

```

## Load all stocks

```{r}
# Read 112 stock files into R environment. Each file is store as an element in a list called stock_file
read_stock_files <- function(directory_path) {
  dir_stock_files <- list.files(directory_path)
  stock_file <- list()
  for (file_name in dir_stock_files) {
    stock_file[[file_name]] <- fread(file.path(directory_path, file_name))
  }
  return(stock_file)
}

stock_file = read_stock_files("individual_book_train")

# NOTE: We won't include original datasets in our submission.

# Instead, to save your time, you may want to load pre-saved RData of stock dataframes.
# They are exactly the same as the ones on Canvas.

# Please uncomment the following readRDS.
# stock_file <- readRDS("RData/stock_file.rds")
```

## Calculate related metrics

```{r Calculate related metrics}
# Calculate WAP1, WAP2, BidAskSpread1, BidAskSpread2 and volume of each transaction
for (i in 1:length(stock_file)) {
  stock_file[[i]] <- stock_file[[i]] |> 
    mutate(WAP1 = (bid_price1 * ask_size1 + ask_price1 * bid_size1) / (bid_size1 + ask_size1)) |> 
    mutate(WAP2 = (bid_price2 * ask_size2 + ask_price2 * bid_size2) / (bid_size2 + ask_size2)) |>  
    mutate(BidAskSpread1 = (ask_price1/bid_price1) - 1) |> 
    mutate(BidAskSpread2 = (ask_price2/bid_price2) - 1) |> 
    mutate(volume = bid_size1 + ask_size1 + bid_size2 + ask_size2)
}
```

## Choosing time_id based on the frequency of transactions

```{r time_id selection, warning=FALSE, message=FALSE}
# Count the number of transactions for each time_id of each stock_id
trans_list <- list()
for (i in 1:length(stock_file)) {
  trans <- stock_file[[i]] |> 
    group_by(time_id, stock_id, ) |> 
    summarise(transactions = n(), .groups = "drop")
  trans_list[[names(stock_file)[i]]] <- trans
}

# Sum up the number of transactions for each time_id across all stock_id
transactions_data <- do.call(rbind, trans_list) |>
  arrange(time_id) |>
  group_by(time_id) |>
  summarize(transactions = sum(transactions))

# Split the distribution of number of transactions into three levels: low, medium and high
breaks <- quantile(transactions_data$transactions, probs = c(0, 1/3, 2/3, 1))
labels <- c("low", "medium", "high")

# Determine the level for each time_id
transactions_data <- transactions_data |> 
  mutate(transaction_level = cut(transactions, breaks = breaks, labels = labels, include.lowest = TRUE))
```

```{r}
ggplot(data = transactions_data, aes(x = transactions)) +
  geom_histogram(bins = 50) +
  labs(x = "Transactions", y = "Frequency") +
  theme_minimal() +
  geom_vline(xintercept = breaks, linetype = "dashed", color = "red")
```

```{r}
# For each level, sample ~ 333 time_ids without replacement so that the total number of time_id is 1000
set.seed(3888)

low_level <- transactions_data |> 
  filter(transaction_level == "low") |> 
  sample_n(1000/3, replace = FALSE)

medium_level <- transactions_data |> 
  filter(transaction_level == "medium") |> 
  sample_n(1000/3, replace = FALSE)

high_level <- transactions_data |> 
  filter(transaction_level == "high") |> 
  sample_n(334, replace = FALSE)

time.id.levels <- bind_rows(low_level, medium_level, high_level)
```

```{r}
# 1000 time_ids vector
time.id.list <- time.id.levels |> arrange(time_id) |> pull(time_id)
```

## Calculate log returns and volatility

```{r log returns and volatility}
# This step is only performed on 1000 time_ids that we have determined before. Calculate log returns for each time_id of a stock. For each time_id, divide into a 30-second bucket to calculate volatility.

comp_vol <- function(x) {return(sqrt(sum(x ^ 2)))}

log_return_list <- list()
vol_list <- list()

for (j in 1:length(stock_file)) {
  
  log_r_list <- list()
  time_IDs <- time.id.list
  stock_id <- unique(stock_file[[j]]$stock_id)

  log_returns <- stock_file[[j]] %>% 
    filter(time_id %in% time_IDs) %>%
    group_by(time_id) %>%
    reframe(
      log_return = log(WAP1[-1] / WAP1[1:(n() - 1)]),
      seconds_in_bucket = seconds_in_bucket[-1]
    ) %>%
    mutate(stock_id = stock_id)

  log_return_list[[names(stock_file)[j]]] <- log_returns
  
  log_r <- log_returns |>  mutate(time_bucket = ceiling(seconds_in_bucket / 30))
  vol <- aggregate(log_return ~ stock_id + time_bucket + time_id, data = log_r, FUN = comp_vol)
  colnames(vol) <- c('stock_id', 'time_bucket', 'time_id', 'volatility')
  
  vol_list[[names(stock_file)[j]]] <- vol
  
}
```

## Feature dataframe

```{r Feature dataframe}
# For each combination of stock_id_time_id e.g. stock_0_16, calculate mean and sd log returns
log_return_df.ls <- list()

for (i in 1:length(stock_file)) {
  
  log_return_df.ls[[names(stock_file)[i]]] <- log_return_list[[i]] |> 
  group_by(time_id, ) |> 
  summarize(
    stock_id = sub("\\.csv", "", names(stock_file)[i]),
    mean_log_return = mean(log_return),
    sd_log_return = sd(log_return)
  )
}

log_return_df <- do.call(rbind, log_return_df.ls) |> arrange(time_id)

```

```{r}
# For each combination of stock_id_time_id e.g. stock_0_16, calculate those features listed below
time_id_df.ls <- list()

for (i in 1:length(stock_file)) {
  
  time_id_df.ls[[names(stock_file)[i]]] <- stock_file[[i]] |> 
    filter(time_id %in% time.id.list) |>
    group_by(time_id) |> 
    summarize(
      stock_id = sub("\\.csv", "", names(stock_file)[i]),
      mean_WAP1 = mean(WAP1),
      med_WAP1 = median(WAP1),
      sd_WAP1 = sd(WAP1),
      mean_WAP2 = mean(WAP2),
      med_WAP2 = median(WAP2),
      sd_WAP2 = sd(WAP2),
      mean_BidAskSpread1 = mean(BidAskSpread1),
      med_BidAskSpread1 = median(BidAskSpread1),
      sd_BidAskSpread1 = sd(BidAskSpread1),
      mean_BidAskSpread2 = mean(BidAskSpread2),
      med_BidAskSpread2 = median(BidAskSpread2),
      sd_BidAskSpread2 = sd(BidAskSpread2),
      mean_volume = mean(volume),
      transactions = n()
    )
}

time_id_df <- do.call(rbind, time_id_df.ls) |> arrange(time_id)
```

```{r}
# Merge the two dataframe above into one feature dataframe called feature_df, 16 features in total
feature_df <- merge(log_return_df, time_id_df, by = c("stock_id", "time_id")) |>
  mutate(stock_time_id = paste0(stock_id, "_", time_id), .before = 1) |>
  arrange(time_id)
```

## K-Means Clustering

```{r K-Means Clustering, warning=FALSE}
# Elbow method to determine the optimal k for K-Means Clustering
wss <- sapply(1:10, function(k) kmeans(feature_df[-(1:3)], centers = k)$tot.withinss)
plot(1:10, wss, type = "b", xlab = "Number of Clusters (k)", ylab = "Within-cluster Sum of Squares (WSS)", main = "Elbow Method for Optimal k")
```

```{r warning=FALSE}
#Apply K-Means clustering with k = 3
set.seed(3888)

cluster_number = 3

normalized_data <- scale(feature_df[-(1:3)])
kmeans <- kmeans(normalized_data, centers = cluster_number)

fviz_cluster(kmeans, data = feature_df[-(1:3)])
```

```{r}
# Add a column with cluster label for each stock_time_id into the feature dataframe
feature_df <- feature_df %>% mutate(cluster = as.character(kmeans$cluster), .before = 1)
```

```{r}
# Sample 20% of data points in each cluster for visualisation
set.seed(3888)
pca_result <- prcomp(feature_df[-c(1:4)], center = TRUE, scale. = TRUE)

feature_df$pca1 <- pca_result$x[, 1]
feature_df$pca2 <- pca_result$x[, 2]

sampled_df <- feature_df %>%
  group_by(cluster) %>%
  sample_frac(0.2, replace = FALSE) %>%
  ungroup()

gg <- ggplot(sampled_df, aes(x = pca1, y = pca2, color = cluster, text = stock_time_id)) +
  geom_point(size = 0.6) +
  labs(x = "PC1", y = "PC2", color = "Cluster") +
  theme_minimal() +
  ggtitle('K-means Clustering Scatter Plot')

ggplotly(gg, tooltip = "text")

```

We summarizds the features by cluster and visualized the distribution of features across clusters using box plots. The goal is to provide a clear understanding of how different features vary across clusters.

```{r summary of clusters}
# Summarize features by cluster
# Remove unnecessary columns and group by 'cluster'
cluster_summary <- feature_df[, c(-2, -3, -4, -21, -22)] %>%
  group_by(cluster) %>%
  summarise(across(everything(), list(mean = mean, median = median), .names = "{fn}_{col}"))

# Convert data to long format for ggplot
data_long <- pivot_longer(feature_df[, c(-2, -3, -4, -21, -22)], cols = -cluster, names_to = "Feature", values_to = "Value")

# Create box plots of feature distributions by cluster
ggplot(data_long, aes(x = cluster, y = Value)) +
  geom_boxplot() +
  facet_wrap(~ Feature, scales = "free") + # Create separate plots for each feature
  theme_minimal() +
  labs(title = "Feature Distributions by Cluster", x = "Cluster", y = "Value")

```

## Cluster

```{r}
# For each cluster, use 50 stock_time_id to build model
set.seed(3888)
eval_df <- feature_df %>%
  group_by(cluster) %>%
  sample_n(50, replace = FALSE) %>%
  ungroup()

# Retrieve log return data for each stock_time_id from the log_return_list that we have created before. Second_in_buckets ranges from 1-600, if a stock_time_id does not have a transaction at any second_in_buckets, log return will be filled with 0.
stock_unique_id <- gsub("[^0-9]", "", names(stock_file))

cluster_feature_ls <- list()

for (cluster_id in 1:3) {

  combined_stock_log_r <- data.frame()
  
  for (i in 1:length(log_return_list)) {

    stock <- eval_df |> 
      filter(cluster == cluster_id & stock_id == paste0('stock_', stock_unique_id[[i]])) |> 
      dplyr::select(stock_id, time_id) |> 
      arrange(stock_id) |> 
      pull(time_id)
    
    stock_log_r <- log_return_list[[i]][log_return_list[[i]]$time_id %in% stock, ]
    
    combined_stock_log_r <- bind_rows(combined_stock_log_r, stock_log_r) |> 
      distinct()
  }
  
  combined_stock_log_r <- combined_stock_log_r %>%
    mutate(stock_time_id = paste0("stock_", stock_id, "_", time_id))
  
  transformed_data <- combined_stock_log_r |> 
    dplyr::select(stock_time_id, seconds_in_bucket, log_return) |> 
    complete(stock_time_id, seconds_in_bucket = 1:600, fill = list(log_return = 0)) |> 
    pivot_wider(names_from = stock_time_id, values_from = log_return)
  
  cluster_feature_ls[[paste0("cluster_", cluster_id)]] <- transformed_data
}

```

```{r}
# Build a feature dataframe for linear regression model, where each stock_time_id will be splitted into 20 buckets of 30 seconds
cluster_feature_wider <- data.frame()

for(i in 1: nrow(eval_df)){
  
  row <- eval_df[i, ]
  
  stock_df <- stock_file[[paste0(row$stock_id, ".csv")]] %>%
    filter(time_id == row$time_id) %>%
    mutate(time_bucket = ceiling(seconds_in_bucket / 30)) %>%
    group_by(time_id, time_bucket) %>%
    summarize(
      mean_WAP1 = mean(WAP1),
      med_WAP1 = median(WAP1),
      sd_WAP1 = sd(WAP1),
      mean_WAP2 = mean(WAP2),
      med_WAP2 = median(WAP2),
      sd_WAP2 = sd(WAP2),
      mean_BidAskSpread1 = mean(BidAskSpread1),
      med_BidAskSpread1 = median(BidAskSpread1),
      sd_BidAskSpread1 = sd(BidAskSpread1),
      mean_BidAskSpread2 = mean(BidAskSpread2),
      med_BidAskSpread2 = median(BidAskSpread2),
      sd_BidAskSpread2 = sd(BidAskSpread2),
      mean_volume = mean(volume),
      transactions = n(),
       .groups = 'drop'
    )
  
  vol <- (vol_list[[paste0(row$stock_id, ".csv")]] %>%
    filter(time_id == row$time_id))[, -3] %>% dplyr::select(-stock_id)
  
  merge_df <- right_join(stock_df, vol, by = "time_bucket") %>%
    arrange(time_bucket) %>%
    mutate(
      stock_time_id = row$stock_time_id,
      stock_id = row$stock_id,
      cluster = row$cluster, 
      .before = 1)
  
  cluster_feature_wider <- rbind(cluster_feature_wider, merge_df)
  
}


cluster_feature_wider <- cluster_feature_wider %>% mutate(across(everything(), ~ifelse(is.na(.), 0, .)))
cluster_feature_wider_cluster <- cluster_feature_wider %>% split(cluster_feature_wider$cluster)

```


## Model Assessment

### Check Stationary of Dataset

```{r warning=FALSE}
for(i in 1:3){
  lag_order <- vector()
  
  for(j in 2: ncol(cluster_feature_ls[[i]])){
    original_adf = tseries::adf.test(unlist(cluster_feature_ls[[i]][, j]), alternative = "stationary")
    if (original_adf$p.value > 0.05) {
      cat("WARNING: not stationary")
    }
  }
}
```

### Build Models

```{r}
# Define the univariate GARCH(1,1) specification without ARMA terms for the mean model
uni_spec <- ugarchspec(
  variance.model = list(
      garchOrder = c(1,1)),  # GARCH(1,1) order
      mean.model = list(
      armaOrder = c(0,0),    # No ARMA terms in the mean model
      include.mean = TRUE    # Include a mean term in the model
  )
)

# Define the univariate ARMA(1,1)-GARCH(1,1) specification with normal distribution for residuals
arma_spec <- ugarchspec(
  variance.model = list(
    model = "sGARCH",       # Use standard GARCH model
    garchOrder = c(1, 1)),  # GARCH(1,1) order
  mean.model = list(
    armaOrder = c(1, 1)),   # ARMA(1,1) order for the mean model
  distribution.model = "norm" # Assume normally distributed residuals
)
```

```{r}
# Function to forecast using a fitted VAR model
forecast_var_model <- function(var_model) {
  # Predict the next 30 time points using the VAR model
  forecast <- predict(var_model, n.ahead = 30)
  # Extract the forecasted values and combine them into a matrix
  pc_forecasts_matrix <- do.call(cbind, lapply(forecast$fcst, function(x) x[, "fcst"]))
  # Transpose the matrix to match the desired format
  return(t(pc_forecasts_matrix))
}

# Function to forecast using a fitted DCC-GARCH model
forecast_dcc_model <- function(dcc_model) {
  # Simulate the next 30 time points using the DCC-GARCH model
  forecast <- dccsim(dcc_model, n.sim = 30, startMethod = "sample", rseed = 3888)@msim$simX[[1]]
  # Transpose the forecast matrix to match the desired format
  pc_forecasts_matrix <- t(forecast)
  # Set the row names of the matrix to the asset names from the DCC model
  rownames(pc_forecasts_matrix) <- dcc_model@model$modeldata$asset.names
  return(pc_forecasts_matrix)
}

# Function to forecast using a fitted ARMA-GARCH model
forecast_arma_model <- function(arma_model) {
  # Simulate future paths for the next 30 time points using the ARMA-GARCH model
  future.path <- fitted(ugarchpath(arma_model, n.sim = 30, m.sim = 1000))
  # Replace NA values in the future path with 0
  future.path[is.na(future.path)] <- 0
  # Calculate the mean of the standard deviations across simulations and return it
  return(mean(apply(future.path, 2, sd)))
}

# Function to perform feature selection using stepwise regression
feature_selection <- function(df){
  # Define the null model with only the intercept
  null_model <- lm(volatility~1, data = df)
  # Define the full model with all potential predictors
  full_model <- lm(volatility~., data = df)

  # Perform stepwise regression using forward selection
  df_forward <- stepAIC(null_model, direction = "forward", 
                        scope = list(lower = null_model, upper = full_model),
                        trace = FALSE)
  
  # Return the selected feature names
  return(labels(terms(df_forward)))
}

```

For each cluster, we performs rolling window forecasting by fitting VAR, DCC-GARCH, Linear Regression, and ARMA-GARCH models to the data in each window. The forecasts and models are stored for later analysis. This approach allows for dynamic and accurate volatility prediction tailored to the unique characteristics of each stock cluster.

```{r model, warning=FALSE}
# Set seed for reproducibility
set.seed(3888)

# Define rolling window parameters
number_of_buckets <- 10
window_size <- number_of_buckets * 30
increment <- 30

# Initialize lists to store forecasts and models for each cluster
forecasts_var.ls <- list()
forecasts_dcc.ls <- list()
forecasts_lr.ls <- list()
forecasts_arma.ls <- list()
var_models.ls <- list()
dcc_models.ls <- list()

# Create a parallel processing cluster
cl <- makeCluster(detectCores() - 1)

# Iterate over each cluster
for (i in 1:length(cluster_feature_ls)) {
  
  cluster <- cluster_feature_ls[[i]]
  cat("Working with cluster", i, "\n")
  
  # Initialize local lists to store forecasts and models for the current cluster
  forecasts_var <- list()
  forecasts_dcc <- list()
  forecasts_lr <- list()
  forecasts_arma <- list()
  var_models <- list()
  dcc_models <- list()
  
  j = 1
  
  # Identify unique stock and time combinations within the cluster
  unique_combinations <- unique(cluster_feature_wider_cluster[[i]][, c("stock_id", "time_id")])
  
  # Perform rolling window forecasting
  for (t in seq(window_size, nrow(cluster) - 30, by = increment)) {
    
    cat("    window from", t - window_size + 1, "to", t, "\n")
    
    # Extract data for the current window
    df <- cluster[(t - window_size + 1):t, -1]
   
    # Fit VAR model and store forecasts and models
    var_model <- vars::VAR(df, p = 4, type = "none")
    forecasts_var[[j]] <- forecast_var_model(var_model)
    var_models[[j]] <- var_model

    # Fit DCC-GARCH model and store forecasts and models
    dcc_model <- dccfit(
      dccspec(uspec = multispec(replicate(ncol(df), uni_spec, simplify = FALSE)),
              dccOrder = c(1, 1), distribution = 'mvnorm'),
      data = df, cluster = cl)

    forecasts_dcc[[j]] <- forecast_dcc_model(dcc_model)
    dcc_models[[j]] <- dcc_model
    
    # Perform feature selection for linear regression
    x = feature_selection(cluster_feature_wider_cluster[[i]][, c(-1, -2, -3, -4, -5)])
    formula <- paste("volatility ~", paste(x, collapse = " + "))
    
    # Iterate over unique stock and time combinations
    for (combo in 1:nrow(unique_combinations)) {
      
      current_stock_id <- unique_combinations$stock_id[combo]
      current_time_id <- unique_combinations$time_id[combo]
      stock_time <- paste(current_stock_id, current_time_id, sep = "_")
    
      # Extract window data and test data for the current combination
      window_data <- cluster_feature_wider_cluster[[i]] %>%
        filter(stock_id == current_stock_id, time_id == current_time_id) %>%
        arrange(time_bucket) %>%
        filter(time_bucket > (t - window_size) / 30 & time_bucket <= (t / 30))
      
      test_data <- cluster_feature_wider_cluster[[i]] %>%
        filter(stock_id == current_stock_id, time_id == current_time_id) %>%
        arrange(time_bucket) %>%
        filter(time_bucket == (t / 30 + 1))
      
      if (nrow(window_data) > 0 & nrow(test_data) > 0) {
        
        # Fit linear regression model and store forecasts
        model <- lm(formula, data = window_data)
        forecasts_lr[[stock_time]][[as.character(j)]] <- predict(model, newdata = test_data)
        
        # Fit ARMA-GARCH model and store forecasts
        arma_garch <- ugarchfit(spec = arma_spec, data = unlist(df[, stock_time]), solver = 'hybrid')
        fspec <- getspec(arma_garch)
        setfixed(fspec) <- as.list(coef(arma_garch))
        forecasts_arma[[stock_time]][[as.character(j)]] <- forecast_arma_model(fspec)
        
      }
    }
    j <- j + 1
    
  }
  
  # Store models and forecasts for the final window
  var_models[[j]] <- vars::VAR(cluster[301:600, -1], p = 4, type = "none")
  dcc_models[[j]] <- dccfit(
    dccspec(uspec = multispec(replicate(ncol(cluster[301:600, -1]), uni_spec, simplify = FALSE)),
            dccOrder = c(1, 1), distribution = 'mvnorm'),
    data = cluster[301:600, -1], cluster = cl)

  cat("\n")
  
  # Store forecasts and models for the current cluster
  forecasts_var.ls[[i]] <- forecasts_var
  forecasts_dcc.ls[[i]] <- forecasts_dcc
  forecasts_lr.ls[[i]] <- forecasts_lr
  forecasts_arma.ls[[i]] <- forecasts_arma
  var_models.ls[[i]] <- var_models
  dcc_models.ls[[i]] <- dcc_models
}

# Stop the parallel processing cluster
stopCluster(cl)

```

### MSE and QLIKE

For each prediction, we calculate the Mean Squared Error (MSE) and Quasi-Likelihood (QLIKE) metrics to assess models.

```{r MSE and QLIKE}
# Initialize lists to store MSE and QLIKE for each model
MSE_var <- list()
QLIKE_var <- list()

MSE_dcc <- list()
QLIKE_dcc <- list()

MSE_lr <- list()
QLIKE_lr <- list()

MSE_arma <- list()
QLIKE_arma <- list()

# Iterate over each cluster (assumed to be 3 clusters)
for (i in 1:3) {
  
  # Filter evaluation data frame for the current cluster
  filtered_df <- eval_df %>% filter(cluster == as.character(i))
  
  # Retrieve forecasts for the current cluster
  prediction_var <- forecasts_var.ls[[i]]
  prediction_dcc <- forecasts_dcc.ls[[i]]
  prediction_lr <- forecasts_lr.ls[[i]]
  prediction_arma <- forecasts_arma.ls[[i]]

  # Iterate over each row in the filtered evaluation data frame
  for (j in 1:nrow(filtered_df)) {
    
    # Initialize lists to store predicted and actual volatilities
    predicted_vol_var <- list()
    predicted_vol_dcc <- list()
    predicted_vol_lr <- list()
    predicted_vol_arma <- list()
    
    actual_vol <- list()
    
    # Get the current row
    row <- filtered_df[j, ]
    index <- 1
    
    # Filter the actual volatility data for the current stock and time
    filtered_vol <- vol_list[[paste0(row$stock_id, ".csv")]] %>% filter(time_id == row$time_id)
    
    # Iterate over the predictions for the current cluster
    for (k in 1:length(prediction_var)) {
      
      # Calculate the corresponding time bucket
      bucket <- ceiling(((k - 1) * 30 + window_size + 1) / 30)
      actual.vol <- filtered_vol %>% filter(time_bucket == bucket) %>% .$volatility
      
      # Check if actual volatility data is available and valid
      if (length(actual.vol) > 0 & all(actual.vol != 0)) {
        # Store the predicted volatilities
        predicted_vol_var[[index]] <- comp_vol(prediction_var[[k]][row$stock_time_id, ])
        predicted_vol_dcc[[index]] <- comp_vol(prediction_dcc[[k]][row$stock_time_id, ])
        predicted_vol_lr[[index]] <- prediction_lr[[row$stock_time_id]][[as.character(k)]]
        predicted_vol_arma[[index]] <- prediction_arma[[row$stock_time_id]][[as.character(k)]]
        
        # Store the actual volatilities
        actual_vol[[index]] <- actual.vol
        index <- index + 1
      }
    }
    
    # Calculate and store MSE for each model
    MSE_var[[as.character(i)]] <- c(MSE_var[[as.character(i)]], 
                                    mean((unlist(actual_vol) - unlist(predicted_vol_var)) ^ 2))
    MSE_dcc[[as.character(i)]] <- c(MSE_dcc[[as.character(i)]],
                                    mean((unlist(actual_vol) - unlist(predicted_vol_dcc)) ^ 2))
    MSE_lr[[as.character(i)]] <- c(MSE_lr[[as.character(i)]], 
                                   mean((unlist(actual_vol) - unlist(predicted_vol_lr)) ^ 2))
    MSE_arma[[as.character(i)]] <- c(MSE_arma[[as.character(i)]], 
                                     mean((unlist(actual_vol) - unlist(predicted_vol_arma)) ^ 2))
    
    # Calculate and store QLIKE for each model
    QLIKE_var[[as.character(i)]] <- c(QLIKE_var[[as.character(i)]], 
                                      mean(unlist(actual_vol) / unlist(predicted_vol_var) - 
                                             log(unlist(actual_vol) /unlist(predicted_vol_var)) - 1))
    
    QLIKE_dcc[[as.character(i)]] <- c(QLIKE_dcc[[as.character(i)]],
                                      mean(unlist(actual_vol) / unlist(predicted_vol_dcc) -
                                             log(unlist(actual_vol) / unlist(predicted_vol_dcc)) - 1))
    
    QLIKE_lr[[as.character(i)]] <- c(QLIKE_lr[[as.character(i)]], 
                                     mean(unlist(actual_vol) / unlist(predicted_vol_lr) -
                                            log(unlist(actual_vol) / unlist(predicted_vol_lr)) - 1))
    
    QLIKE_arma[[as.character(i)]] <- c(QLIKE_arma[[as.character(i)]], 
                                       mean(unlist(actual_vol) / unlist(predicted_vol_arma) -
                                              log(unlist(actual_vol) / unlist(predicted_vol_arma))- 1))
  }
}

```

```{r}
par(mar=c(2, 2, 2, 2), mfrow=c(4, 3))

models <- list(
  "Vector Auto-Regression" = MSE_var,
  "DCC-GARCH" = MSE_dcc,
  "Linear Regression" = MSE_lr,
  "ARMA-GARCH" = MSE_arma
)

for (model_name in names(models)) {
  for (i in 1:3) {
    boxplot(models[[model_name]][[as.character(i)]], horizontal = TRUE,
            main = paste("MSE of Cluster", i, "\nUsing", model_name), cex.main = 0.9)
  }
}

par(mfrow=c(1, 1))

```

```{r}
par(mar=c(2, 2, 2, 2), mfrow=c(4, 3))

models <- list(
  "Vector Auto-Regression" = QLIKE_var,
  "DCC-GARCH" = QLIKE_dcc,
  "Linear Regression" = QLIKE_lr,
  "ARMA-GARCH" = QLIKE_arma
)

for (model_name in names(models)) {
  for (i in 1:3) {
    boxplot(models[[model_name]][[as.character(i)]], horizontal = TRUE,
            main = paste("QLIKE of Cluster", i, "\nUsing", model_name), cex.main = 0.9)
  }
}

par(mfrow=c(1, 1))
```

### Example Plot

```{r example plot}
stock_name_ex = "stock_10"
time_id_ex = 14126

cluster_ex <- eval_df %>% filter(stock_id == stock_name_ex & time_id == time_id_ex)

forecasts_var <- forecasts_var.ls[[as.integer(cluster_ex$cluster)]]
forecasts_dcc <- forecasts_dcc.ls[[as.integer(cluster_ex$cluster)]]

predicted_logr_var <- data.frame()
predicted_logr_dcc <- data.frame()

predicted_vol_var <- data.frame()
predicted_vol_dcc <- data.frame()

for(i in 1: length(forecasts_var)){
  predicted_logr_var <- rbind(predicted_logr_var, data.frame(
    seconds_in_bucket = seq((i+number_of_buckets-1)*30+1, (i+number_of_buckets)*30),
    log_return = forecasts_var[[i]][paste(stock_name_ex, time_id_ex, sep="_"), ],
    source = "Vector Auto-Regression Predicted"
  ))
  
  predicted_vol_var <- rbind(predicted_vol_var, data.frame(
    time_bucket = i+number_of_buckets,
    volatility = comp_vol(forecasts_var[[i]][paste(stock_name_ex, time_id_ex, sep="_"), ]),
    source = "Vector Auto-Regression Predicted"
  ))
}


for(i in 1: length(forecasts_dcc)){
  
  predicted_logr_dcc <- rbind(predicted_logr_dcc, data.frame(
    seconds_in_bucket = seq((i+number_of_buckets-1)*30+1, (i+number_of_buckets)*30),
    log_return = forecasts_dcc[[i]][paste(stock_name_ex, time_id_ex, sep="_"), ],
    source = "DCC-GARCH Predicted"
  ))

  predicted_vol_dcc <- rbind(predicted_vol_dcc, data.frame(
    time_bucket = i+number_of_buckets,
    volatility = comp_vol(forecasts_dcc[[i]][paste(stock_name_ex, time_id_ex, sep="_"), ]),
    source = "DCC-GARCH Predicted"
  ))
}

```

```{r}
actual_logr <- merge(data.frame(seconds_in_bucket = seq(1, 600)), 
                     log_return_list[[paste(stock_name_ex, ".csv", sep="")]] %>% 
                       filter(time_id == time_id_ex) %>%
                       dplyr::select(seconds_in_bucket, log_return), 
                     by = "seconds_in_bucket", all.x = TRUE) %>%
  mutate(source = "Realised Log Returns")
actual_logr$log_return[is.na(actual_logr$log_return)] <- 0


actual_vol <-  merge(data.frame(
  time_bucket = seq(1, 20)), 
  vol_list[[paste(stock_name_ex, ".csv", sep="")]] %>% 
    filter(time_id == time_id_ex) %>%
    dplyr::select(time_bucket, volatility), 
  by = "time_bucket", all.x = TRUE) %>%
  mutate(source = "Realised Volatility")
actual_vol$volatility[is.na(actual_vol$volatility)] <- 0
```


```{r}
var_model <- var_models.ls[[as.integer(cluster_ex$cluster)]][[11]]
dcc_model <- dcc_models.ls[[as.integer(cluster_ex$cluster)]][[11]]

forecasts_var <- data.frame(forecast_var_model(var_model))
forecasts_dcc <- data.frame(forecast_dcc_model(dcc_model))

future_logr_var <- data.frame(
  seconds_in_bucket = seq(601, 630),
  log_return = unlist(forecasts_var[paste(stock_name_ex, time_id_ex, sep="_"), ]),
  source = "Vector Auto-Regression Future"
)

future_logr_dcc <- data.frame(
  seconds_in_bucket = seq(601, 630),
  log_return = unlist(forecasts_dcc[paste(stock_name_ex, time_id_ex, sep="_"), ]),
  source = "DCC-GARCH Future"
)

future_vol_var <- data.frame(
  time_bucket = c(20, 21),
  volatility = c(actual_vol$volatility[20], comp_vol(future_logr_var$log_return)),
  source = "Vector Auto-Regression Future")

future_vol_dcc <- data.frame(
  time_bucket = c(20, 21),
  volatility = c(actual_vol$volatility[20], comp_vol(future_logr_dcc$log_return)),
  source = "DCC-GARCH Future")

```

```{r}
logr_var_plt <- rbind(actual_logr, predicted_logr_var, future_logr_var) %>%
  ggplot(aes(x = seconds_in_bucket, y = log_return, color = source)) +
  geom_line() +
  scale_color_manual(values = c("Realised Log Returns" = "grey", 
                                "Vector Auto-Regression Predicted" = "blue", 
                                "Vector Auto-Regression Future" = "red")) +
  labs(title = paste0("Log Returns and Predictions for ", stock_name_ex, " at time ", time_id_ex),
       x = "Seconds",
       y = "Log Return",
       color = "Source") +
  theme_minimal()

logr_dcc_plt <- rbind(actual_logr, predicted_logr_dcc, future_logr_dcc) %>%
  ggplot(aes(x = seconds_in_bucket, y = log_return, color = source)) +
  geom_line() +
  scale_color_manual(values = c("Realised Log Returns" = "grey", 
                                "DCC-GARCH Predicted" = "blue",
                                "DCC-GARCH Future" = "red")) +
  labs(title = paste0("Log Returns and Predictions for ", stock_name_ex, " at time ", time_id_ex),
       x = "Seconds",
       y = "Log Return",
       color = "Source") +
  theme_minimal()


grid.arrange(logr_var_plt, logr_dcc_plt, nrow = 2)
```

```{r}
vol_var_plt <- rbind(actual_vol, predicted_vol_var, future_vol_var) %>%
  ggplot(aes(x = time_bucket, y = volatility, color = source)) +
    geom_line() + 
    geom_point() +
    scale_color_manual(values = c("Realised Volatility" = "grey", 
                                  "Vector Auto-Regression Predicted" = "blue",
                                  "Vector Auto-Regression Future" = "red")) +
    labs(title = "Volatility and Predictions for Example Stock for future 30 seconds",
         x = "Time Bucket",
         y = "Volatility",
         color = "Source") +
    theme_minimal()

vol_dcc_plt <- rbind(actual_vol, predicted_vol_dcc, future_vol_dcc) %>%
  ggplot(aes(x = time_bucket, y = volatility, color = source)) +
    geom_line() +
    geom_point() +
    scale_color_manual(values = c("Realised Volatility" = "grey", 
                                  "DCC-GARCH Predicted" = "blue",
                                  "DCC-GARCH Future" = "red")) +
    labs(title = "Volatility and Predictions for Example Stock for future 30 seconds",
         x = "Time Bucket",
         y = "Volatility",
         color = "Source") +
    theme_minimal()


grid.arrange(vol_var_plt, vol_dcc_plt, nrow = 2)
```

## Stability

### Bootstrap Method

Generate bootstrap samples, then test the models on these

If the performance does not significantly decrease on the samples, this indicates model stability


```{r bootstrap}
# First, create a function that returns bootstrap samples of the input sample, using block bootstrap
# Block bootstrap is a bootstrap method commonly used for time series data
# Return value is a list

block_bootstrap <- function(x, block_size = 30, n_samples = 1) {
  n_blocks <- nrow(x) %/% block_size
  if (n_blocks * block_size < nrow(x)) {n_blocks <- n_blocks + 1}

  # calculate mean and sdev for each block for each stock
  block.means <- list()
  block.sdevs <- list()

  for (i in 1 : ncol(x)) {
    block.means.i <- list()
    block.sdevs.i <- list()

    for (j in 1 : n_blocks) {
      upper_limit <- j * block_size
      lower_limit <- upper_limit - block_size + 1
      if (j == n_blocks) {upper_limit <- nrow(x)}

      block.means.i[[j]] <- mean(unlist(x[lower_limit:upper_limit,i]))
      block.sdevs.i[[j]] <- sd(unlist(x[lower_limit:upper_limit,i]))
    }

    block.means[[i]] <- block.means.i
    block.sdevs[[i]] <- block.sdevs.i
  }

  # create the bootstrap samples

  # one list for each sample, then one sublist for each column
  bootstrap.samples <- list()

  for (k in 1 : n_samples) {
    bootstrap.samples.k <- list()
    for (i in 1 : ncol(x)) {
      sample.i <- list()
      for (j in 1 : n_blocks) {
        sample.i[[j]] <- rnorm(block_size, mean = block.means[[i]][[j]], block.sdevs[[i]][[j]])
      }
      bootstrap.samples.k[[i]] <- unlist(sample.i)
    }
    bootstrap.samples[[k]] <- t(do.call(rbind, bootstrap.samples.k))
  }

  return(bootstrap.samples)
}

```

```{r}
# Test the models on a bootstrap sample
set.seed(3888)

# bs stands for bootstrap
bs.forecasts_var.ls <- list()
bs.forecasts_dcc.ls <- list()

# Create the bootstrap samples
bs.cluster_feature_ls <- list()
for (i in 1:length(cluster_feature_ls)) {
  bs.cluster_feature_ls[[i]] <- block_bootstrap(cluster_feature_ls[[i]])
}

# randomly select one window
t <- sample(seq(window_size, nrow(cluster) - 30, by = 30), 1)

cl <- makeCluster(detectCores() - 1)

# Loop through each cluster
for (i in 1:length(bs.cluster_feature_ls)) {

  cluster <- bs.cluster_feature_ls[[i]][[1]]
  
  df <- cluster[(t-window_size+1):t, -1]

  var_model <- vars::VAR(df, p = 4, type = "none")
  forecasts_var <- forecast_var_model(var_model)
  rownames(forecasts_var) <- colnames(cluster_feature_ls[[i]][-1])

  arma_model <- dccfit(
    dccspec(uspec = multispec(replicate(ncol(df), uni_spec, simplify = FALSE)),
            dccOrder = c(1, 1), distribution = 'mvnorm'),
    data = df, cluster = cl)
  forecasts_dcc <- forecast_dcc_model(arma_model)
  rownames(forecasts_dcc) <- colnames(cluster_feature_ls[[i]][-1])

  bs.forecasts_var.ls[[i]] <- forecasts_var
  bs.forecasts_dcc.ls[[i]] <- forecasts_dcc
}

stopCluster(cl)


```

### MSE and QLIKE on bootstrap samples

```{r}
# The similarity between the performance on the bootstrap samples and the original will indicate high stability

BS_MSE_var <- list()
BS_QLIKE_var <- list()
BS_MSE_dcc <- list()
BS_QLIKE_dcc <- list()

for(i in 1: 3){

   filtered_df <- eval_df %>% filter(cluster == as.character(i))

   prediction_var <- bs.forecasts_var.ls[[i]]
   prediction_dcc <- bs.forecasts_dcc.ls[[i]]

   for(j in 1: nrow(filtered_df)){

    row <- filtered_df[j, ]
    
    filtered_vol <- vol_list[[paste0(row$stock_id, ".csv")]] %>% filter(time_id == row$time_id)
    
    actual_vol <- filtered_vol %>% filter(time_bucket == (t / 30 + 1)) %>% .$volatility
    
    predicted_vol_var <- comp_vol(prediction_var[row$stock_time_id, ])
    predicted_vol_dcc <- comp_vol(prediction_dcc[row$stock_time_id, ])

    BS_MSE_var[[as.character(i)]] <- c(BS_MSE_var[[as.character(i)]],
                                    (actual_vol - predicted_vol_var) ^ 2)
    BS_MSE_dcc[[as.character(i)]] <- c(BS_MSE_dcc[[as.character(i)]],
                                    (actual_vol - predicted_vol_dcc) ^ 2)
    
    BS_QLIKE_var[[as.character(i)]] <- c(BS_QLIKE_var[[as.character(i)]], 
                                         actual_vol / predicted_vol_var -
                                                log(actual_vol / predicted_vol_var) - 1)
    BS_QLIKE_dcc[[as.character(i)]] <- c(BS_QLIKE_dcc[[as.character(i)]], 
                                          actual_vol / predicted_vol_dcc -
                                                log(actual_vol / predicted_vol_dcc) - 1)

   }

}


```

```{r}
models <- list(
  "Vector Auto-Regression" = list(MSE_var, BS_MSE_var),
  "DCC-GARCH" = list(MSE_dcc, BS_MSE_dcc)
)

par(mar=c(2, 2, 2, 2), mfrow=c(2, 3))


for (model_name in names(models)) {
  for (i in 1:3) {
    boxplot(models[[model_name]][[1]][[as.character(i)]],
            models[[model_name]][[2]][[as.character(i)]], horizontal = TRUE,
            main = paste("MSE of Cluster", i, "\nUsing", model_name), 
            names = c("Original", "Bootstrap"),
            cex.main = 0.9)
  }
}

par(mfrow=c(1, 1))
```

```{r}
models <- list(
  "Vector Auto-Regression" = list(QLIKE_var, BS_QLIKE_var),
  "DCC-GARCH" = list(QLIKE_dcc, BS_QLIKE_dcc)
)

par(mar=c(2, 2, 2, 2), mfrow=c(2, 3))


for (model_name in names(models)) {
  for (i in 1:3) {
    boxplot(models[[model_name]][[1]][[as.character(i)]],
            models[[model_name]][[2]][[as.character(i)]], horizontal = TRUE,
            main = paste("QLIKE of Cluster", i, "\nUsing", model_name), 
            names = c("Original", "Bootstrap"),
            cex.main = 0.9)
  }
}

par(mfrow=c(1, 1))
```

### Stability


```{r stability}
# A function evaluates the stability of VAR and DCC-GARCH models by computing the eigenvalues of their companion matrices across rolling windows.

stability_df.fun <- function(stock_name_ex, time_id_ex, model.ls, model_string) {
  # Filter evaluation data frame for the specified stock and time ID
  cluster_ex <- eval_df %>% filter(stock_id == stock_name_ex & time_id == time_id_ex)
  # Initialize data frames to store eigenvalues
  eigen_windows_alpha.values.df <- as.data.frame(matrix(nrow = 50))
  
  # Initialize eigenvalue data frames based on model type
  if (model_string == "var") {
    eigen_windows.values.df <- as.data.frame(matrix(nrow = 200))
  } else if (model_string == "dcc") {
    eigen_windows.values.df <- as.data.frame(matrix(nrow = 50))
  }
  
  # Get the list of models for the specified cluster
  cluster_model.ls <- model.ls[[as.numeric(cluster_ex$cluster)]]
  number_of_features <- 50
  number_of_observations <- 300 # Each window
  
  # Iterate over each window model in the cluster
  for (window.index in seq(cluster_model.ls)) {
    window_model <- cluster_model.ls[[window.index]]
    
    # Get coefficients for the specified stock and time ID
    id <- paste0(stock_name_ex, "_", time_id_ex)
    
    if (model_string == "var") {
      # Calculate eigenvalues for VAR model
      estimate <- data.frame(coefficients(window_model)[[id]])$Estimate
      companion_matrix <- array(data = estimate, dim = c(200, 200))
      eigen_vec_val <- abs(eigen(as.matrix(companion_matrix))$values)
      
      etemp.df <- as.data.frame(eigen_vec_val)
      colnames(etemp.df) <- paste0("eigen_val", window.index)
      eigen_windows.values.df <- cbind(eigen_windows.values.df, etemp.df)
      
    } else if (model_string == "dcc") {
      # Calculate eigenvalues for DCC-GARCH model
      estimate <- data.frame(coef(window_model))
      alpha1 <- estimate[grepl(".alpha1$", rownames(estimate)), ]
      beta1 <- estimate[grepl(".beta1$", rownames(estimate)), ]
      
      # Eigenvalues for beta1 parameter
      companion_matrix_beta1 <- array(data = beta1, dim = c(50, 50))
      eigen_vec_val_beta1 <- abs(eigen(as.matrix(companion_matrix_beta1))$values)
      etemp.df_beta1 <- as.data.frame(eigen_vec_val_beta1)
      colnames(etemp.df_beta1) <- paste0("eigen_val_beta_", window.index)
      eigen_windows.values.df <- cbind(eigen_windows.values.df, etemp.df_beta1)
      
      # Eigenvalues for alpha1 parameter
      companion_matrix_alpha1 <- array(data = alpha1, dim = c(50, 50))
      eigen_vec_val_alpha1 <- abs(eigen(as.matrix(companion_matrix_alpha1))$values)
      etemp.df_alpha1 <- as.data.frame(eigen_vec_val_alpha1)
      colnames(etemp.df_alpha1) <- paste0("eigen_val_alpha_", window.index)
      eigen_windows_alpha.values.df <- cbind(eigen_windows_alpha.values.df, etemp.df_alpha1)
    }
  }
  
  # Prepare the result to return
  return_result <- list()
  return_result[[1]] <- eigen_windows.values.df[, -1] # Remove the initial empty column
  return_result[[2]] <- eigen_windows_alpha.values.df[, -1] # Remove the initial empty column
  return(return_result)
}

```


```{r}
# var var_models.ls
# example for stability: stock 10, 14126
stock_name_ex = "stock_10"
time_id_ex = 14126
result <- stability_df.fun(stock_name_ex, time_id_ex, var_models.ls, "var")
eigen_windows.df <- result[[1]]

```

```{r}
stable_check <- c()
for (row.index in seq(nrow(eigen_windows.df))){
  for (col.index in seq(length(eigen_windows.df))){
    if (eigen_windows.df[row.index, col.index] < 1){
      stable_check <- c(stable_check, TRUE)
    }else{
      stable_check <- c(stable_check, FALSE)
    }
  }
}

```

```{r}
cat(paste0("The var model is stable: ",(sum(stable_check == TRUE)/length(stable_check))*100, 
           "% of the modulus of eigenvalue is less than 1."))
```

```{r}
eigen_windows_long.df <- eigen_windows.df %>%
  mutate(lag = 1:n()) %>%
  pivot_longer(cols = starts_with("eigen_val"), 
               names_to = "window", 
               values_to = "eigenvalue") %>%
  mutate(window = factor(window, levels = paste0("eigen_val", 1:11), labels = paste0("window", 1:11)))

p <- ggplot(eigen_windows_long.df, aes(x = lag, y = eigenvalue, color = window)) + 
  geom_point() + 
  xlab("lags") + 
  ylab("eigenvalue") +
  theme_minimal()

ggplotly(p)
```


```{r}
# dcc_models.ls - DCC - GARCH
# example for stability: stock 10, 14126
stock_name_ex = "stock_10"
time_id_ex = 14126
result <- stability_df.fun(stock_name_ex, time_id_ex, dcc_models.ls, "dcc")
beta.df <- result[[1]]
alpha.df <- result[[2]]

```

```{r}
stable_check <- c()
for (row.index in seq(nrow(beta.df))){
  for (col.index in seq(length(beta.df))){
    if (beta.df[row.index, col.index] < 1){
      stable_check <- c(stable_check, TRUE)
    }else{
      stable_check <- c(stable_check, FALSE)
    }
  }
}

```

```{r}
cat(paste0("The Multivariate DCC-GARCH model - GARCH term is stable: ",(sum(stable_check == TRUE)/length(stable_check))*100, 
           "% of the modulus of eigenvalue is less than 1."))
```

```{r}
beta_long.df <- beta.df %>%
  mutate(lag = 1:n()) %>%
  pivot_longer(cols = starts_with("eigen_val_beta_"), 
               names_to = "window", 
               values_to = "eigenvalue") %>%
  mutate(window = factor(window, levels = paste0("eigen_val_beta_", 1:11), labels = paste0("window", 1:11)))

p <- ggplot(beta_long.df, aes(x = lag, y = eigenvalue, color = window)) + 
  geom_point() + 
  xlab("lags") + 
  ylab("eigenvalue") +
  theme_minimal()

ggplotly(p)
```

```{r}
stable_check <- c()
for (row.index in 1:nrow(alpha.df)){
  col.index <- 1
  while (col.index < length(alpha.df)){
    
    if ((alpha.df[row.index, col.index]) < 1){
      stable_check <- c(stable_check, TRUE)
    }else{
      stable_check <- c(stable_check, FALSE)
    }
    col.index = col.index + 1
  }
}

cat(paste0("The Multivariate DCC-GARCH model - ARCH term is stable: ",(sum(stable_check == TRUE)/length(stable_check))*100, 
           "% of the modulus of eigenvalue is less than 1."))
```

```{r}
alpha.long.df <- alpha.df %>%
  mutate(lag = 1:n()) %>%
  pivot_longer(cols = starts_with("eigen_val_alpha"), 
               names_to = "window", 
               values_to = "eigenvalue") %>%
  mutate(window = factor(window, levels = paste0("eigen_val_alpha_", 1:11), labels = paste0("window", 1:11)))

p <- ggplot(alpha.long.df, aes(x = lag, y = eigenvalue, color = window)) + 
  geom_point() + 
  xlab("lags") + 
  ylab("eigenvalue") +
  theme_minimal()

ggplotly(p)
```

